---
title: "sgRNA Epigenetics - sgRNA Sequences Tables"
author: "Gianluca Pegoraro"
date: today
date-format: long
format: 
  html:
      self-contained: true
      code-fold: true
      code-tools: true
      code-link: true
editor: source
---

## README

This script is used to generate a table of sequences for the sgRNAs present in the sgEpigenetic library from Synthego + the ones present in the sgGuin custom library.



Plates HT001453, HT001456, HT001459 (All plates 3 of the screen) were unusable because most of the cells from columns 17 to 24 in these plates were lost. We identified the issue with the Bluewasher program for cell fixation that caused this issue and corrected it. As a result, a new batch of plates 1 - 3 was spotted:

## Setup

Load packages.

```{r}
#| label: load-packages
#| message: false

library(tidyverse)
library(fs)
library(paint)
```

## sgRNA input and wrangling

Read the sgRNA layout information provided by Synthego for the Epigenetics Library. 3 sgRNA oligos per well. 3 plates.

```{r}
#| label: read-epi-gene-list

epi_gene_tbl <- 
  dir_ls(path = "reformat_metadata/Synthego",
         glob = "*qc.csv") |>
  read_csv(
    id = "file_name",
    col_names = c("mother_well",
                  "gene_symbol",
                  "oligo_seq"),
    col_types = "ccc"
  ) |>
  select(gene_symbol, oligo_seq) |>
  group_by(gene_symbol) |>
  mutate(oligo_id = seq(1:n())) |>
  pivot_wider(id_cols = gene_symbol,
              names_prefix = "sgrna_oligo_",
              names_from = oligo_id, 
              values_from = oligo_seq) |>
  ungroup()

paint(epi_gene_tbl)
```

Read the sgRNA layout information provided by Synthego for the Guin Library. 3 sgRNA oligos per well. 1 plate.

```{r}
#| label: read-guin-gene-list

guin_gene_tbl <- read_csv("reformat_metadata/Synthego/guin_library.csv") |>
  select(gene_symbol = gene,
         oligo_seq = customer_sequence) |>
  mutate(oligo_seq = str_replace_all(oligo_seq, "\\*", ""),
         oligo_seq = str_replace_all(oligo_seq, " \\+ Synthego modified EZ Scaffold", "")) |>
  group_by(gene_symbol) |>
  mutate(oligo_id = seq(1:n())) |>
  filter(oligo_id <= 3) |>
  pivot_wider(id_cols = gene_symbol,
              names_prefix = "sgrna_oligo_",
              names_from = oligo_id, 
              values_from = oligo_seq) |>
  ungroup()

glimpse(guin_gene_tbl)
```

Concatenate Epigenetics gene table with Guin gene table.

```{r}
#| label: contatenate-gene-lists

gene_tbl <- bind_rows(epi_gene_tbl, guin_gene_tbl)

glimpse(gene_tbl)
```

Write the gene table to a file.
```{r}
#| label: write-gene-table

write_csv(gene_tbl, "knitr_output/sgRNA_sequences_table.csv")
```

Document the information about the analysis session

```{r}
sessionInfo()
```
